---
title: 'Comparing public health outcomes of interventions'
teaching: 45 # teaching time in minutes
exercises: 30 # exercise time in minutes

---

:::::::::::::::::::::::::::::::::::::: questions 

- How can I quantify the effect of an intervention?

 
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Compare intervention scenarios

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites
+ Complete tutorials [Simulating transmission](../episodes/simulating-transmission.md) and [Modelling interventions](../episodes/modelling-interventions.md)

Learners should familiarise themselves with following concept dependencies before working through this tutorial: 

**Outbreak response** : [Intervention types](https://www.cdc.gov/nonpharmaceutical-interventions/).
:::::::::::::::::::::::::::::::::


## Introduction

In this tutorial we will compare intervention scenarios against each other. To quantify the effect of the intervention we need to compare our intervention scenario to a counter factual scenario. The *counter factual* is the scenario in which nothing changes, often referred to as the 'do nothing' scenario. The counter factual scenario may include no interventions, or if we are investigating the potential impact of an additional intervention in the later stages of an outbreak there may be existing interventions in place. 

We must also decide what our *outcome of interest* is to make comparisons between intervention and counter factual scenarios. The outcome of interest can be:

+ a model outcome, e.g. number of infections or hospitalisations,
+ a metric such as the epidemic peak time or size,
+ a measure that uses the model outcomes such as QALY/DALYs.


:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

In this tutorial we introduce the concept of the counter factual and how to compare scenarios (counter factual versus intervention) against each other. 

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

## Vacamole model

The Vacamole model is a deterministic model based on a system of ODEs in [Ainslie et al. 2022]( https://doi.org/10.2807/1560-7917.ES.2022.27.44.2101090) to describe the effect of vaccination on COVID-19 dynamics. The model consists of 11 compartments, individuals are classed as one of the following:

+ susceptible, $S$,
+ partial vaccination ($V_1$), fully vaccination ($V_2$),
+ exposed, $E$ and exposed while vaccinated, $E_V$,
+ infectious, $I$ and infectious while vaccinated, $I_V$,
+ hospitalised, $H$ and hospitalised while vaccinated, $H_V$,
+ dead, $D$,
+ recovered, $R$.

The diagram below describes the flow of individuals through the different compartments. 

```{r, echo = FALSE, message = FALSE}
DiagrammeR::grViz("digraph{
  # graph statement
  #################
  graph [layout = dot,
         rankdir = LR,
         overlap = true,
         fontsize = 10]

  # nodes
  #######
  node [shape = square,
       fixedsize = true
       width = 1.3]

       S
       E
       Ev [label = 'E@_{V}', style = filled, fillcolour = 'gray']
       I
       Iv [label = 'I@_{V}', style = filled, fillcolour = 'gray']
       H
       Hv [label = 'H@_{V}', style = filled, fillcolour = 'gray']
       D
       R
       V1 [label = 'V@_{1}', style = filled, fillcolour = 'gray']
       V2 [label = 'V@_{2}', style = filled, fillcolour = 'gray']


  # edges
  #######
  S -> E [label = ' infection (&beta;) ']
  S -> V1 [label = ' vaccination (&nu;1)']
  V1 -> E [label = ' infection (&beta;)']
  V1 -> V2 [label = ' vaccination\n(second dose) (&nu;2)']
  V2 -> Ev [label = ' infection (&beta;)']
  Ev -> Iv [label = ' onset of \ninfectiousness (&alpha;) ']
  E -> I [label = ' onset of \ninfectiousness (&alpha;) ']
  I -> H [label = ' hospitalisation (&eta;)']
  Iv -> Hv [label = ' hospitalisation (&eta;@_{V})']
  I -> D [label = ' death (&omega;)']
  I -> R [label = ' recovery (&gamma;)']
  Iv -> D [label = ' death (&omega;@_{V})']
  Iv -> R [label = ' recovery (&gamma;)']
  Hv -> D [label = ' death (&omega;@_{V})']
  Hv -> R [label = ' recovery (&gamma;)']
  H -> D [label = ' death (&omega;)']
  H -> R [label = ' recovery (&gamma;)']

  subgraph {
  rank = same; S; V1;V2;
  }
}")
```


::::::::::::::::::::::::::::::::::::: challenge

## Running a counterfactual scenario using the Vacamole model 

1. Run the model with the default parameter values for the UK population assuming that :

+ 1 in a million individual are infectious (and not vaccinated) at the start of the simulation
+ The contact matrix for the United Kingdom has age bins:
  + age between 0 and 20 years,
  + age between 20 and 40,
  + 40 years and over.
+ There is no vaccination scheme in place

2. Using the output, plot the number of deaths through time


::::::::::::::::: hint

### Vaccination code 

To run the model with no vaccination in place we can *either* create two vaccination objects (one for each dose) using `vaccination()` with the time start, time end and vaccination rate all set to 0, or we can use the `no_vaccination()` function to create a vaccination object for two doses with all values set to 0.

```{r, eval = FALSE}
no_vaccination <- no_vaccination(population = uk_population, doses = 2)
```
::::::::::::::::::::::

::::::::::::::::: hint

### HINT : Running the model with default parameter values

We can run the Vacamole model with [default parameter values](https://epiverse-trace.github.io/epidemics/articles/vacamole.html#model-epidemic-using-vacamole) by just specifying the population object and number of time steps to run the model for: 


```{r, eval = FALSE}
output <- model_vacamole_cpp(
  population = uk_population,
  vaccination = no_vaccination,
  time_end = 300
)
```

::::::::::::::::::::::



::::::::::::::::: solution

### SOLUTION

1.  Run the model

```{r}
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  survey = polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# extract demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# prepare initial conditions
initial_i <- 1e-6

initial_conditions <- c(
  S = 1 - initial_i,
  V1 = 0, V2 = 0,
  E = 0, EV = 0,
  I = initial_i, IV = 0,
  H = 0, HV = 0, D = 0, R = 0
)

initial_conditions <- rbind(
  initial_conditions,
  initial_conditions,
  initial_conditions
)
rownames(initial_conditions) <- rownames(contact_matrix)

# prepare population object
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)

no_vaccination <- no_vaccination(population = uk_population, doses = 2)

# run model
output <- model_vacamole_cpp(
  population = uk_population,
  vaccination = no_vaccination,
  time_end = 300
)
```

2. Plot the number of deaths through time

```{r}
ggplot(output[output$compartment == "dead", ]) +
  geom_line(
    aes(time, value, colour = demography_group),
    linewidth = 1
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    labels = rownames(contact_matrix),
    name = "Age group"
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Simulation time (days)",
    y = "Individuals"
  ) +
  theme(
    legend.position = "top"
  ) +
  theme_bw(
    base_size = 15
  )
```



:::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::::::::::::::



## Comparing scenarios

*Coming soon*



## Challenge : Ebola outbreak analysis 

*Coming soon*




::::::::::::::::::::::::::::::::::::: keypoints 

- The counter factual scenario must be defined to make comparisons

::::::::::::::::::::::::::::::::::::::::::::::::
