---
title: 'Introduction'
teaching: 10
exercises: 2
editor_options: 
  chunk_output_type: console
---

:::::::::::::::::::::::::::::::::::::: questions 

- Why to use R packages for Outbreak analytics?
- What can we do to analyse our outbreak data?
- How can I start doing Outbreak analytics with R?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain our vision on the need of outbreak analytics R packages. 
- Share our strategy to create R packages into an outbreak analytics pipeline.
- Define our plan to start your learning path in outbreak analytics with R.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

List (and hyperlink) the lessons/packages which need to be covered before this lesson

```r
install_version(
  package = "EpiNow2", version = "1.4.0", 
  repos = "http://cran.us.r-project.org"
)
```

:::::::::::::::::::::::::::::::::


## Introduction

_write about the Reproduction number (in a motivational way)_

Packages are handy tools to reuse code, maintenance, less error prone data analysis steps.

The `{EpiNow2}` provide us with a a three-steps solution for this task!

```{r,warning=FALSE}
library(EpiNow2)
```


## First, get your data

A data frame of observation

```{r}
example_confirmed
```

## Then, set the parameters

```{r}
incubation_period <- dist_spec(
  mean = 4, sd = 2,
  max = 20, distribution = "gamma"
)

reporting_delay <- dist_spec(
  mean = convert_to_logmean(2, 1), 
  sd = convert_to_logsd(2, 1),
  max = 10, distribution = "lognormal"
)

generation_time <- dist_spec(
  mean = 3.6, sd = 3.1,
  max = 20, distribution = "lognormal"
)
```

## Let's calculate R!

```{r,message=FALSE,warning=FALSE,cache=TRUE}
epinow_estimates <- epinow(
  # cases
  reported_cases = example_confirmed[1:60],
  # delays
  generation_time = generation_time_opts(generation_time),
  delays = delay_opts(
    incubation_period + reporting_delay
  ),
  # computation
  stan = stan_opts(
    cores = 4, samples = 1000, chains = 2,
    control = list(adapt_delta = 0.99)
  )
)
```

```{r}
base::plot(epinow_estimates)
```


## The problem!

However, doing this in real life is not as easy as this example!

Data analysis involves dealing with inputs problems. 

- Read your linelist
- Clean your linelist
- Validate your linelist
- Read parameters from literature

Also you can usethis R outputs as inputs for other tasks.

- Forecast cases
- Estimate severity
- Simulate transmission scenarios
- Compare interventions

## What can we do?

At Epiverse-TRACE we are creating packages that complement the current landscape filling gaps of epi-specific challenges in response to outbreaks.

![An overview of the tutorial task to cover.](https://epiverse-trace.github.io/task_pipeline-minimal.svg)


## How can we start?

In first set of episodes we are going to deal with each of these task previous to the _Quantify transmission_ task. These preliminary task are the __Early tasks__. Then we are going to get deeper into the _Quantify transmission_ task, which is within the __Middle tasks__, and later ones in the pipeline called and __Late tasks__.

::::::::::::::::::::::::::::::::::::: keypoints 

- Our vision is to have pipelines of R packages for outbreak analytics.
- Our strategy is to create interconnected tasks to get public health relevant outputs.
- Our plan is to introduce about package solutions and theory bits for each of the tasks in the outbreak analytics pipeline.

::::::::::::::::::::::::::::::::::::::::::::::::

