---
title: 'Modelling interventions'
teaching: 45 # teaching time in minutes
exercises: 30 # exercise time in minutes

---

```{r setup, echo= FALSE, message = FALSE, warning = FALSE}
require(ggplot2)
library(epidemics)
```

:::::::::::::::::::::::::::::::::::::: questions 

- How do I investigate the effect of interventions on disease trajectories? 


::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Add pharmaceutical and non-pharmaceutical interventions to an {epidemics} model

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites
+ Complete tutorial [Simulating transmission](../episodes/simulating-transmission.md)

Learners should familiarise themselves with following concept dependencies before working through this tutorial: 

**Outbreak response** : [Intervention types](https://www.cdc.gov/nonpharmaceutical-interventions/).
:::::::::::::::::::::::::::::::::


## Introduction

Mathematical models can be used to generate trajectories of disease spread under the implementation of interventions at different stages of an outbreak. These predictions can be used to make decisions on what interventions could be implemented to slow down the spread of diseases. 

We can assume interventions in mathematical models reduce the values of relevant parameters e.g. reduce transmissibility while in place. Or it may be appropriate to assume individuals are classified into a new disease state, e.g. once vaccinated we assume individuals are no longer susceptible to infection and therefore move to a vaccinated state. In this tutorial, we will introduce how to include three different interventions in model of COVID-19 transmission. 

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

In this tutorial different types of intervention and how they can be modelled are introduced. Learners should be able to understand the underlying mechanism of these interventions (e.g. reduce contact rate) as well as how to implement the code to include such interventions.

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

## Non-pharmaceutical interventions

[Non-pharmaceutical interventions](../learners/reference.md#NPIs) (NPIs) are measures put in place to reduce transmission that do not include the administration of drugs or vaccinations. NPIs aim reduce contact between infectious and susceptible individuals. For example, washing hands, wearing masks and closures of school and workplaces.

We will investigate the effect of interventions on a COVID-19 outbreak using an SEIR model (`model_default_cpp()` in the R package `{epidemics}`). We will set $R_0 = 2.7$, latent period or preinfectious period $= 4$ and the infectious_period $= 5.5$ (parameters adapted from [Davies et al. (2020)](https://doi.org/10.1016/S2468-2667(20)30133-X)). We load a contact matrix with age bins 0-18, 18-65, 65 years and older using `{socialmixr}` and assume that one in every 1 million in each age group is infectious at the start of the epidemic.

```{r model_setup, echo = TRUE, message = FALSE}
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 15, 65),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# initial conditions: one in every 1 million is infected
initial_i <- 1e-6
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

# build for all age groups
initial_conditions <- matrix(
  rep(initial_conditions, dim(contact_matrix)[1]),
  ncol = 5, byrow = T,
)
rownames(initial_conditions) <- rownames(contact_matrix)

# prepare the population to model as affected by the epidemic
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)
```

#### Effect of school closures on COVID-19 spread

The first NPI we will consider is the effect of school closures on reducing the number of individuals infectious with COVID-19 through time. We assume that a school closure will reduce the frequency of contacts within and between different age groups. We assume that school closures will reduce the contacts between school aged children (aged 0-15) by 0.5, and will cause a small reduction (0.01) in the contacts between adults (aged 15 and over). 

To include an intervention in our model we must create an `intervention` object. The inputs are the name of the intervention (`name`), the type of intervention (`contacts` or `rate`), the start time (`time_begin`), the end time (`time_end`) and the reduction (`reduction`). The values of the reduction matrix are specified in the same order as the age groups in the contact matrix. 

```{r}
rownames(contact_matrix)
```

Therefore, we specify ` reduction = matrix(c(0.5, 0.01, 0.01))`. We assume that the school closures start on day 50 and are in place for a further 100 days. Therefore our intervention object is : 

```{r intervention}
close_schools <- intervention(
  name = "School closure",
  type = "contacts",
  time_begin = 50,
  time_end = 50 + 100,
  reduction = matrix(c(0.5, 0.01, 0.01))
)
```

::::::::::::::::::::::::::::::::::::: callout
### Effect of interventions on contacts

In `epidemics`, the contact matrix is scaled down by proportions for the period in which the intervention is in place. To understand how the reduction is calculated within the model functions, consider a contact matrix for two age groups with equal number of contacts:

```{r echo = FALSE}
reduction <- matrix(c(0.5, 0.1))
contact_matrix_example <- matrix(c(1, 1, 1, 1), nrow = 2)
contact_matrix_example
```

If the reduction is 50% in group 1 and 10% in group 2, the contact matrix during the intervention will be:

```{r echo = FALSE}
contact_matrix_example[1, ] <- contact_matrix_example[1, ] * (1 - reduction[1])
contact_matrix_example[, 1] <- contact_matrix_example[, 1] * (1 - reduction[1])
contact_matrix_example[2, ] <- contact_matrix_example[2, ] * (1 - reduction[2])
contact_matrix_example[, 2] <- contact_matrix_example[, 2] * (1 - reduction[2])
contact_matrix_example
```

The contacts within group 1 are reduced by 50% twice to accommodate for a 50% reduction in outgoing and incoming contacts ($1\times 0.5 \times 0.5 = 0.25$). Similarly, the contacts within group 2 are reduced by 10% twice. The contacts between group 1 and group 2 are reduced by 50% and then by 10% ($1 \times 0.5 \times 0.9= 0.45$). 

::::::::::::::::::::::::::::::::::::::::::::::::

Using transmissibility $= 2.7/5.5$ (remember that [transmissibility = $R_0$/ infectious period](../episodes/simulating-transmission.md#the-basic-reproduction-number-r_0)), infectiousness rate $1/= 4$ and the recovery rate $= 1/5.5$ we run the model with` intervention = list(contacts = close_schools)` as follows :

```{r school}
output_school <- model_default_cpp(
  population = uk_population,
  transmissibility = 2.7 / 5.5,
  infectiousness_rate = 1.0 / 4.0,
  recovery_rate = 1.0 / 5.5,
  intervention = list(contacts = close_schools),
  time_end = 300, increment = 1.0
)
```


To be able to see the effect of our intervention, we also run the model where there is no intervention, combine the two outputs into one data frame and then plot the output. Here we plot the total number of infectious individuals in all age groups using `ggplot2::stat_summary()`:

```{r baseline, echo = TRUE, fig.width = 10}
# run baseline simulation with no intervention
output_baseline <- model_default_cpp(
  population = uk_population,
  transmissibility = 2.7 / 5.5,
  infectiousness_rate = 1.0 / 4.0,
  recovery_rate = 1.0 / 5.5,
  time_end = 300, increment = 1.0
)

# create intervention_type column for plotting
output_school$intervention_type <- "school closure"
output_baseline$intervention_type <- "baseline"
output <- rbind(output_school, output_baseline)

ggplot(data = output[output$compartment == "infectious", ]) +
  aes(
    x = time,
    y = value,
    color = intervention_type,
    linetype = intervention_type
  ) +
  stat_summary(
    fun = "sum",
    geom = "line",
    linewidth = 1
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Simulation time (days)",
    y = "Individuals"
  ) +
  theme_bw(
    base_size = 15
  ) +
  geom_vline(
    xintercept = c(close_schools$time_begin, close_schools$time_end),
    colour = "black",
    linetype = "dashed",
    linewidth = 0.2
  ) +
  annotate(
    geom = "text",
    label = "Schools closed",
    colour = "black",
    x = (close_schools$time_end - close_schools$time_begin) / 2 +
      close_schools$time_begin,
    y = 10,
    angle = 0,
    vjust = "outward"
  )
```
We see that with the intervention in place, the infection still spreads through the population, though the peak number of infectious individuals is smaller than the baseline with no intervention in place (solid line).



#### Effect of mask wearing on COVID-19 spread

We can model the effect of other NPIs as reducing the value of relevant parameters. For example, we want to investigate the effect of mask wearing on the number of individuals infectious with COVID-19 through time. 

We expect that mask wearing will reduce an individual's infectiousness. As we are using a population based model, we cannot make changes to individual behaviour and so assume that the transmissibility $\beta$ is reduced by a proportion due to mask wearing in the population. We specify this proportion, $\theta$ as product of the proportion wearing masks multiplied by the proportion reduction in transmissibility (adapted from [Li et al. 2020](https://doi.org/10.1371/journal.pone.0237691))

We create an intervention object with `type = rate` and `reduction = 0.161`. Using parameters adapted from [Li et al. 2020](https://doi.org/10.1371/journal.pone.0237691) we have proportion wearing masks = coverage $\times$ availability = $0.54 \times 0.525 = 0.2835$, proportion reduction in transmissibility = $0.575$. Therefore, $\theta = 0.2835 \times 0.575 = 0.163$. We assume that the mask wearing mandate starts at day 40 and is in place for 200 days.

```{r masks}
mask_mandate <- intervention(
  name = "mask mandate",
  type = "rate",
  time_begin = 40,
  time_end = 40 + 200,
  reduction = 0.163
)
```

To implement this intervention on the parameter $\beta$, we specify `intervention = list(beta = mask_mandate)`.

```{r output_masks}
output_masks <- model_default_cpp(
  population = uk_population,
  transmissibility = 2.7 / 5.5,
  infectiousness_rate = 1.0 / 4.0,
  recovery_rate = 1.0 / 5.5,
  intervention = list(transmissibility = mask_mandate),
  time_end = 300, increment = 1.0
)
```


```{r plot_masks, echo = TRUE, message = FALSE, fig.width = 10}
# create intervention_type column for plotting
output_masks$intervention_type <- "mask mandate"
output_baseline$intervention_type <- "baseline"
output <- rbind(output_masks, output_baseline)

ggplot(data = output[output$compartment == "infectious", ]) +
  aes(
    x = time,
    y = value,
    color = intervention_type,
    linetype = intervention_type
  ) +
  stat_summary(
    fun = "sum",
    geom = "line",
    linewidth = 1
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Simulation time (days)",
    y = "Individuals"
  ) +
  theme_bw(
    base_size = 15
  ) +
  geom_vline(
    xintercept = c(mask_mandate$time_begin, mask_mandate$time_end),
    colour = "black",
    linetype = "dashed",
    linewidth = 0.2
  ) +
  annotate(
    geom = "text",
    label = "Mask mandate",
    colour = "black",
    x = (mask_mandate$time_end - mask_mandate$time_begin) / 2 +
      mask_mandate$time_begin,
    y = 10,
    angle = 0,
    vjust = "outward"
  )
```

::::::::::::::::::::::::::::::::::::: callout
### Intervention types

There are two intervention types for `model_default_cpp()`. Rate interventions on model parameters (`transmissibillity` $\beta$, `infectiousness_rate` $\sigma$ and `recovery_rate` $\gamma$) and contact matrix reductions `contacts`.

To implement both contact and rate interventions in the same simulation they must be passed as a list e.g. `intervention = list(transmissibility = mask_mandate, contacts = close_schools)`. But if there are multiple interventions that target contact rates, these must be passed as one `contacts` input. See the [vignette on modelling overlapping interventions](https://epiverse-trace.github.io/epidemics/articles/multiple_interventions.html) for more detail. 

::::::::::::::::::::::::::::::::::::::::::::::::


## Pharmaceutical interventions

Pharmaceutical interventions (PIs) are measures such as vaccination and mass treatment programs. In the previous section, we assumed that interventions reduced the value of parameter values while the intervention was in place. In the case of vaccination, we assume that after the intervention individuals are no longer susceptible and should be classified into a different disease state. Therefore, we specify the rate at which individuals are vaccinated and track the number of vaccinated individuals through time. 

The diagram below shows the SEIRV model implemented using `model_default_cpp()` where susceptible individuals are vaccinated and then move to the $V$ class.

```{r diagram_SEIRV, echo = FALSE, message = FALSE}
DiagrammeR::grViz("digraph {

  # graph statement
  #################
  graph [layout = dot,
         rankdir = LR,
         overlap = true,
         fontsize = 10]

  # nodes
  #######
  node [shape = square,
       fixedsize = true
       width = 1.3]

       S
       E
       I
       R
       V

  # edges
  #######
  S -> E [label = ' infection (&beta;)']
  S -> V [label = ' vaccination (&nu;)']
  E -> I [label = ' onset of \ninfectiousness (&alpha;)']
  I -> R [label = ' recovery (&gamma;)']

}")
```



The equations describing this model are as follows: 

$$
\begin{aligned}
\frac{dS_i}{dt} & = - \beta S_i \sum_j C_{i,j} I_j -\nu_{t} S_i \\
\frac{dE_i}{dt} &= \beta S_i\sum_j C_{i,j} I_j - \alpha E_i \\
\frac{dI_i}{dt} &= \alpha E_i - \gamma I_i \\
\frac{dR_i}{dt} &=\gamma I_i \\
\frac{dV_i}{dt} & =\nu_{i,t} S_i\\
\end{aligned}
$$
Individuals are vaccinated at an age group ($i$) specific time dependent ($t$) vaccination rate ($\nu_{i,t}$). The SEIR components of these equations are described in the tutorial [simulating transmission](../episodes/simulating-transmission.md#simulating-disease-spread). 

To explore the effect of vaccination we need to create a vaccination object to pass as an input into `model_default_cpp()` that includes an age groups specific vaccination rate `nu` and age group specific start and end times of the vaccination program (`time_begin` and `time_end`). 

Here we will assume all age groups are vaccinated at the same rate 0.01 and that the vaccination program starts on day 40 and is in place for 150 days.

```{r vaccinate}
# prepare a vaccination object
vaccinate <- vaccination(
  name = "vaccinate all",
  time_begin = matrix(40, nrow(contact_matrix)),
  time_end = matrix(40 + 150, nrow(contact_matrix)),
  nu = matrix(c(0.01, 0.01, 0.01))
)
```

We pass our vaccination object using `vaccination = vaccinate`:

```{r output_vaccinate}
output_vaccinate <- model_default_cpp(
  population = uk_population,
  transmissibility = 2.7 / 5.5,
  infectiousness_rate = 1.0 / 4.0,
  recovery_rate = 1.0 / 5.5,
  vaccination = vaccinate,
  time_end = 300, increment = 1.0
)
```


::::::::::::::::::::::::::::::::::::: challenge 

## Compare interventions

Plot the three interventions vaccination, school closure and mask mandate and the baseline simulation on one plot. Which intervention reduces the peak number of infectious individuals the most?


:::::::::::::::::::::::: solution 

## Output

```{r plot_vaccinate, echo = TRUE, message = FALSE, fig.width = 10}
# create intervention_type column for plotting
output_vaccinate$intervention_type <- "vaccination"
output <- rbind(output_school, output_masks, output_vaccinate, output_baseline)

ggplot(data = output[output$compartment == "infectious", ]) +
  aes(
    x = time,
    y = value,
    color = intervention_type,
    linetype = intervention_type
  ) +
  stat_summary(
    fun = "sum",
    geom = "line",
    linewidth = 1
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Simulation time (days)",
    y = "Individuals"
  ) +
  theme_bw(
    base_size = 15
  )
```

From the plot we see that the peak number of total number of infectious individuals when vaccination is in place is much lower compared to school closures and mask wearing interventions. 

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::




## Summary

Different types of intervention can be implemented using mathematical modelling. Modelling interventions requires assumptions of which model parameters are affected (e.g. contact matrices, transmissibility), by what magnitude and and what times in the simulation of an outbreak. 

The next step is to quantify the effect of an interventions. If you are interested in learning how to compare interventions, please complete the tutorial [Comparing public health outcomes of interventions](../episodes/compare-interventions.md). 

::::::::::::::::::::::::::::::::::::: keypoints 

- The effect of NPIs can be modelled as reducing contact rates between age groups or reducing the transmissibility of infection
- Vaccination can be modelled by assuming individuals move to a different disease state $V$

::::::::::::::::::::::::::::::::::::::::::::::::
