---
title: 'Quantifying transmissibility'
teaching: 30
exercises: 0
---

```{r setup, echo= FALSE, warning = FALSE, message=FALSE}
require(EpiNow2);require(ggplot2); require(incidence2)
require(webshot)
webshot::install_phantomjs(force = TRUE)
```

:::::::::::::::::::::::::::::::::::::: questions 

- How can I estimate the effective reproduction number from case data?
- How do I perform short-term forecasts of infections from case data

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn how to estimate transmissibility features from case data using the R package `EpiNow2`

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

This tutorial has the following concept dependencies: 

**Statistics** : probability distributions, principle of Bayesian analysis. 

**Epidemic theory** : Effective reproduction number.

:::::::::::::::::::::::::::::::::


## Introduction

In an outbreak situation, data are often available on reported infections but not the exact date of infection. To create short term forecasts of future infections and to estimate the effective reproduction number (time-varying reproduction number) we must use estimation methods which account for delays between time of infection and time of case report. 

In the next lessons we will focus on how to implement the functions in `{EpiNow2}` to estimate transmissibility features of case data from an outbreak. We will not cover the theoretical background of the models or inference framework, for details on these concepts see the  [vignette](https://epiforecasts.io/EpiNow2/dev/articles/estimate_infections.html).


::::::::::::::::::::::::::::::::::::: callout
### Bayesian inference

The R package `EpiNow2` uses a Bayesian inference framework to estimate reproduction numbers and infection times based on reporting dates.

In Bayesian inference, we use prior knowledge (prior distributions) with data (in a likelihood function) to find the posterior probability.

<p class="text-center" style="background-color: white">Posterior probability $\propto$ likelihood $\times$ prior probability
</p>

::::::::::::::::::::::::::::::::::::::::::::::::


The first step is to install the package :

```{r, eval = FALSE}
install.packages("EpiNow2")
library(EpiNow2)
```



:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

This lesson illustrates the usage of `epinow()` to estimate the time-varying reproduction number and infection times. Learners should understand the necessary inputs to the model and the limitations of the model output. 

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::


## Delay distributions and case data 
### Case data

To illustrate the functions of `EpiNow2` we will use outbreak data of the start of the COVID-19 pandemic from the United Kingdom. The data are available in the R package `{incidence2}`. 

```{r}
library(incidence2)
head(covidregionaldataUK)
```

To use the data, we must format the data to have two columns:

+ `date` : the date,
+ `confirm` : number of confirmed cases on that date.

```{r}
cases <- covidregionaldataUK[, c("date", "cases_new")]
cases <- aggregate(x = cases$cases_new,
                     FUN = sum,
                     by = list(cases$date),
                     na.rm = TRUE)
colnames(cases) <- c("date", "confirm")
```


There are case data available for `r dim(cases)[1]` days, but in an outbreak situation it is likely we would only have access to the beginning of this data set. Therefore we assume we only have the first 90 days of this data. 

```{r echo = FALSE}
ggplot(cases[1:90,], aes(x = date, y = confirm)) +
  geom_col()
```



### Delay distributions 
We must specify a delay distribution from time of infection until time reported. This delay distribution can consist of multiple types of delays/processes. A typical delay distribution from time of infection to case notification may consist of :

<p class="text-center" style="background-color: aliceblue">**time from infection to symptom onset** (the latent period) + **time from symptom onset to case notification** (the reporting time)
.</p>

The delay distribution for each of these processes can either estimated from data or obtained from the literature. The distributions can also either be **fixed** (i.e. have no uncertainty) or *variable** (i.e. have associated uncertainty).

#### Latent period distribution 

The distribution of latent period can usually be obtained from the literature. The package `{epiparameter}` contains a library of epidemiological parameters for different diseases obtained from the literature. 

We specify a (fixed) gamma distribution with mean $\mu = 4$ and standard deviation $\sigma^2= 2$ (shape = $4$, scale = $1$) using the function `dist_spec()` as follows:

```{r}
latent_period_fixed <- dist_spec(
  mean = 4, sd = 2, 
  max = 20, distribution = "gamma"
)
latent_period_fixed 
```

The argument `max` is the maximum value the distribution can take, in this example 20 days. 

::::::::::::::::::::::::::::::::::::: callout
### Why a gamma distrubution? 

The latent period has to be positive in value. Therefore we must specific a distribution in `dist_spec` which is for positive values only. 

`dist_spec()` supports lognormal and gamma distribution, both of which are distributions for positive values only. 


::::::::::::::::::::::::::::::::::::::::::::::::



####  Including distribution uncertainty

To specify a **variable** distribution, we include uncertainty around the mean $\mu$ and standard deviation $\sigma^2$ of our gamma distribution. This is done as follows, if our latent period distribution has a mean $\mu$ and standard deviation $\sigma^2$. 

Then we assume the mean ($\mu$) follows a Normal distribution with standard deviation $\sigma_{\mu}^2$:

$$\mbox{Normal}(\mu,\sigma_{\mu}^2)$$

and a standard deviation ($\sigma^2$) follows a Normal distribution with standard deviation $\sigma_{\sigma^2}^2$:

$$\mbox{Normal}(\sigma^2,\sigma_{\sigma^2}^2).$$

We specify this using `dist_spec` with the additional arguments `mean_sd` ($\sigma_{\mu}^2$) and `sd_sd` ($\sigma_{\sigma^2}^2$).

```{r}
latent_period_variable <- dist_spec(
  mean = 4, sd = 2,
  mean_sd = 0.5, sd_sd = 0.5,
  max = 20, distribution = "gamma"
  )
latent_period_variable 
```



####  Reporting delays

After the latent period, there will be an additional delay of time from symptom onset to case notification : the reporting delay. We can specify this as a fixed or variable distribution, or estimate a distribution from data. 

When specifying a distribution, it is useful to visualise the probability density to see the peak and spread of the lognormal distribution. We can use the functions `convert_to_logmean()` and `convert_to_logsd()` to  convert the mean and standard deviation of a normal distribution to that of a lognormal distribution. 

```{r}
log_mean <- convert_to_logmean(2, 1)
log_sd <- convert_to_logsd(2, 1) 
x <- seq(from = 0,to = 10 ,length = 1000)
df <- data.frame(x = x, density = dlnorm(x, meanlog = log_mean, sdlog = log_sd))
ggplot(df) +
  geom_line(aes(x, density))
```

Then we can specify a fixed or variable distribution using `dist_spec()`: 

```{r}
reporting_delay_variable <- dist_spec(
  mean = log_mean, sd = log_sd,
  mean_sd = 0.5, sd_sd = 0.5,
  max = 10, distribution = "lognormal"
  )
```


If data is available on the time between symptom onset and reporting, we can use the function `estimate_delay()` to estimate a lognormal distribution from a vector of delays. The code below illustrates how to use `estimate_delay()` with synthetic delay data. 

```{r, eval = FALSE }
delay_data <- rlnorm(500, log(5), 1) # synthetic delay data 
reporting_delay <- estimate_delay(
  delay_data, samples = 1000, 
  bootstraps = 10
  )
```


####  Generation time

We also must specify a distribution for the generation time. Here we will use a Lognormal distribution with mean 3.6 and standard deviation 3.1 ([Ganyani et al. 2020] (https://doi.org/10.2807/1560-7917.ES.2020.25.17.2000257)).


```{r}
generation_time_variable <- dist_spec(
  mean = 3.6, sd = 3.1,
  mean_sd = 0.5, sd_sd = 0.5,
  max = 20, distribution = "lognormal"
)
```


## Finding estimates

The function `epinow()` is a wrapper for the function `estimate_infections()` used to estimate cases by date of infection. As well as the case data and delay distributions, there are numerous other inputs for see `EpiNow2::?epinow()` for more detail.

One optional input is to specify a lognormal prior for the effective reproduction number $R_t$ at the start of the outbreak.

```{r, eval = FALSE}
rt_log_mean <- convert_to_logmean(1, 0.5)
rt_log_sd <- convert_to_logsd(1, 0.5) 
rt = rt_opts(prior = list(mean = rt_log_mean, sd = rt_log_sd ))
```



```{r, echo = FALSE}

rt_log_mean <- convert_to_logmean(2, 1)
rt_log_sd <- convert_to_logsd(2, 1) 

latent_period_fixed <- dist_spec(
  mean = 4, sd = 2, 
  max = 20, distribution = "gamma"
)

log_mean <- convert_to_logmean(2, 1)
log_sd <- convert_to_logsd(2, 1) 
reporting_delay_fixed <- dist_spec(
  mean = log_mean, sd = log_sd,
  max = 10, distribution = "lognormal"
)

generation_time_fixed <- dist_spec(
  mean = 3.6, sd = 3.1,
  max = 20, distribution = "lognormal"
)
```

*Note : in the code below fixed distributions are used instead of variable. This is to speed up computation time. Using variable distributions accounts for additional uncertainty is recommended.*

```{r, message = FALSE, eval = TRUE}
estimates <- epinow(
  reported_cases = cases[1:90,],
  generation_time = generation_time_opts(generation_time_fixed),
  delays = delay_opts(latent_period_fixed + reporting_delay_fixed),
  rt = rt_opts(prior = list(mean = rt_log_mean, sd = rt_log_sd))
)
```


::::::::::::::::::::::::::::::::::::: callout
### The Effective Reproduction Number, $R_t$ 

The basic reproduction number, $R_0$, is the average number of cases caused by one infectious individual in a entirely susceptible population. 

But in an ongoing outbreak, the population is not entirely susceptible. Therefore we are more interested in the value of the **effective reproduction number**, $R_t$, the average number of cases caused by one infectious individual in the population at time $t$.

The effective reproduction number is related to $R_0$ as follows, 

$$R_t = s R_0.$$

where $s=S(t)/N$, the proportion of the population susceptible to infection.


::::::::::::::::::::::::::::::::::::::::::::::::

We can visualise the estimates of the effective reproduction number. The estimates are split into three categories:

+ **Estimate** : utilises all data,

+ **Estimate based on partial data** : estimates that are based less data are are therfore more uncertain,

+ **Forecast** : forecasts into the future. 


```{r}
plot(estimates)

```


We can also extract a summary of the key transmissibility features at the latest date in the data:

```{r}
summary(estimates)
```

## Summary

`EpiNow2` can be used to estimate infections and create forecasts using case data. The quality of the estimates relies on the quality often data and appropriate delay distributions. In the next lesson we will learn how to make regional estimates and investigate some of the additional inference options available in `EpiNow2`. 

::::::::::::::::::::::::::::::::::::: keypoints 

- Uncertainty can be accounted for in delay distributions

::::::::::::::::::::::::::::::::::::::::::::::::
