---
title: 'Simulate transmission chains'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to simulate transmission chains?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Create a short term projection using branching process with `{epichains}`.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites

Learners should familiarise themselves with following concept dependencies before working through this tutorial: 

**Statistics**: probability distributions. 

**Epidemic theory**: Reproduction number.

:::::::::::::::::::::::::::::::::

## questions

## Introduction

<!-- what we know -->

<!-- what we dont know -->

<!-- what we are going to do -->

We are going to use `{purrr}`

```{r,message=FALSE,warning=FALSE}
library(epichains)
library(tidyverse)
```

## Simulation of uncontrolled outbreaks

simulation of uncontrolled outbreaks

- branching process
    + To assess the effect of individual variation on disease outbreaks, we analyse a branching process model with negative binomial offspring distribution

- estimate tree parameters using branching process
    + estimate size
    + estimate length

with estimated parameter we then can reconstruct tranmission chains to forecast and simulate interventions

### branching process

- probability generating function (pgf) of the offspring distribution 

## Get the offspring distribution

from MERS estimates in estimate episode

```{r}
mers_offspring <- 
  c(
    mean = 0.60, 
    dispersion = 0.02
  )

mers_offspring
```

instead of the offspring distribution
we can also define a poisson process
using the lambda distribution parameter
lambda = rate = mean


## Get generation time

```{r}
library(epiparameter)

serial_inverval <- epidist_db(
  disease = "mers",
  epi_dist = "serial",
  single_epidist = T
)

serial_inverval
```

Similar to the `{cfr}` interface for `{epiparameter}`

```r
function(x) generate(x = serial_inverval,times = x)
```

## Simulate a single chain

```{r}
simulate_chains(
  # simulation controls
  index_cases = 5,
  statistic = "size",
  # offspring
  offspring_dist = rnbinom,
  mu = mers_offspring["mean"],
  size = mers_offspring["dispersion"],
  # generation
  generation_time = function(x) generate(x = serial_inverval,times = x)
) %>% 
  as_tibble()
```

the output dataframe collects
infectees as observation unit
each infectee has a sim_id
the infector sim_id of each infectee is in infector_id
infectee got infected in generation and time
as part of the simulation_n chain

## Simulate multiple chains

```{r}
# Set seed for random number generator
set.seed(33)
# Number of simulation runs
number_chains <- 1e3
# Number of initial cases
initial_cases <- 1

```


```{r}
simulated_chains_list <- 
  map(
    .x = seq_len(number_chains), 
    .f = function(sim) {
      simulate_chains(
        # simulation controls
        index_cases = initial_cases,
        statistic = "size",
        # offspring
        offspring_dist = rnbinom,
        mu = mers_offspring["mean"],
        size = mers_offspring["dispersion"],
        # generation
        generation_time = function(x) generate(x = serial_inverval,times = x)
      ) %>% 
        mutate(simulation_n = sim)
    }
  )

sim_chains_bind <- bind_rows(simulated_chains_list)

sim_chains_bind
```

```{r,echo=FALSE}
#### get simulated chain with more infections --------------------------------

chain_to_observe <- sim_chains_bind %>% 
  as_tibble() %>% 
  count(simulation_n,sort = TRUE) %>% 
  filter(n == max(n)) %>% 
  pull(simulation_n)
```

get simulated chain `r chain_to_observe` with more infections

get epichain summary

```{r}
#### get epichain summary ----------------------------------------------------

simulated_chains_list %>% 
  pluck(chain_to_observe)
```

the `NA` includes the infector of the index case, for that reason (known) cases

infectee-infector data frame

```{r}
#### infector-infectee data frame --------------------------------------------

simulated_chains_list %>% 
  pluck(chain_to_observe) %>% 
  as.data.frame()
```


## Visualize multiple chains

```{r}
# daily aggregate of cases
simulated_chains_day <- 
  sim_chains_bind %>% 
  as_tibble() %>% 
  mutate(simulation_n = as_factor(simulation_n)) %>% 
  # Daily number of cases for each simulation
  mutate(day = ceiling(time)) %>%
  count(simulation_n, day, name = "cases") %>% 
  # Cumulative number of cases for each simulation
  group_by(simulation_n) %>% 
  mutate(cases_cumsum = cumsum(cases)) %>% 
  ungroup()

# Summarise the chain length and size 
sim_chains_max <- 
  simulated_chains_day %>%
  group_by(simulation_n) %>% 
  summarise(
    # length
    day_max = max(day),
    # size
    cases_total = max(cases_cumsum)
  ) %>% 
  ungroup()

# Visualize transmission chains by cumulative cases
ggplot() +
  geom_line(
    data = simulated_chains_day,
    mapping = aes(
      x = day, 
      y = cases_cumsum, 
      group = simulation_n,
      color = simulation_n
    ),
    show.legend = FALSE
  ) +
  geom_point(
    data = sim_chains_max,
    mapping = aes(
      x = day_max,
      y = cases_total,
      group = simulation_n,
      color = simulation_n
    ),
    show.legend = FALSE
  ) +
  # xlim(0,150) +
  # ylim(0,300) +
  NULL

# Visualize transmission chains by daily cases
ggplot() +
  geom_line(
    data = simulated_chains_day,
    mapping = aes(
      x = day, 
      y = cases, 
      group = simulation_n,
      color = simulation_n
    ),
    show.legend = FALSE
  )
```


## Challenges

:::::::::::::::::::::::::: challenge

Evaluate for ebola with know data from previous outbreaks

```{r}
epidist_db(
  disease = "ebola",
  epi_dist = "offspring",
  single_epidist = T
)

epidist_db(
  disease = "ebola",
  epi_dist = "serial interval",
  single_epidist = T
)
```


:::::::::::::::::::::::::

:::::::::::::::::::::::::: challenge

from paper
Christian Althaus
<https://www.thelancet.com/journals/laninf/article/PIIS1473-3099(15)70135-0/fulltext>

data from Fayre et al on ebola

```{r}
# Number of individuals in the trees
n <- 152
# Number of secondary cases for all individuals
c1 <- c(1,2,2,5,14,1,4,4,1,3,3,8,2,1,1,4,9,9,1,1,17,
        2,1,1,1,4,3,3,4,2,5,1,2,2,1,9,1,3,1,2,1,1,2)
c0 <- c(c1,rep(0,n-length(c1)))

c0 %>% 
  enframe() %>% 
  ggplot(aes(value)) +
  geom_histogram()
```


::::::::::: hint

...

:::::::::::

::::::::::: solution

remarkable even with R0 less than 1
with potentially explosive outbreaks

the observed variation in individual infectiousness in Ebola
means that although the probability of extinction is high,
new index cases also have the potential for explosive
regrowth of the epidemic.

:::::::::::

:::::::::::::::::::::::::::::

::::::::::::::::::::::::: testimonial

you can use early linelist data as seed cases to make short term projection of cases

take a look to vignette

<https://epiverse-trace.github.io/epichains/articles/projecting_incidence.html>

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- Use `{epichains}`

::::::::::::::::::::::::::::::::::::::::::::::::

